\ourtitle

### 2020-05-12 paper/article.tex
Add examples to the paper

\section{Introduction}
Programming languages are sometimes described as data plus code. However, it is a known fact there is some overlap between the two. While a big chunk of data most programs consume is ingested through some sort of \textit{IO} operation (like reading from a keyboard, a file or a socket), and another chunk is generated dynamically by algorithms at runtime, there is a third chunk: smaller is amount but maybe just as important: constants (or literals) inserted into the code by the programmer.

@@

When we cross into the territory of DSLs \textit{(Domain Specific Languages)}, there is a multitude of custom syntax for the most varied data types. In this paper we will discuss a proposed solution to the problem of describind musical notes, and more important, musical arrangements and melodies as data values that can be integrated into a programming language.

@@

The most straightforward way for generating such timed events is to let the user control the timing manually: setting both properties (start and duration) explicitly for each event, and maybe doing so with regular function calls, thus avoiding the need for extra syntax even. This is the approach used by some of the existing languages in this space, such as \textit{SonicPi}. But this approach is also cumbersome and prone to mistakes though, as one could guess from the following usecase: if the user wants to change one of the earlier events' timing, he would have to manually update the timings of all of the following events.

\begin{lstlisting}[caption={Example of a hypothetical imperative API for creating events},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
play( 0, 100, 'A' );
play( 100, 50, 'B' );
play( 150, 200, 'C' );
\end{lstlisting}

Instead, we'll take a look at an alternative approach based on custom operators and syntax, each with their own semantic and syntax that enable an expressive way to define those events. We can look at this the same way languages implement other primitive values: most often we can declare static strings, not by manually setting each character in memory, but by writing the string wrapped in special characters. For musical events, we will be doing the same, exploring a way to define them in code, as \textit{musical literals}.

@@

The second requirement, \textbf{laziness}, refers to a concept that is familiar in functional programming languages: values are generated when we need them, not earlier. In our case, this implies that a musical sequence could be potentially infinite (like an infinite repetition of some arrangement). If playing this music live, the musician could determine when to stop this arrangement sooner or later.

Given these two requirements, we can conclude we cannot generate all music events at the start and then play them in order.

@@

@@

\paragraph*{Operators}
Operators are special operations defined at the syntactic level that allow \textit{music} to be composed in different ways, such as concatenated, parallelized or repeated.

\begin{description}
    \item[Concatenation] \verb|Music1 Music2 ... MusicN|
    \item[Parallel] \verb'Music1 | Music2 | ... | MusicN'
    \item[Repetition] \verb'Music * Integer'
\end{description}

It is also useful to estabilish that while most operators work on sequences of musical events, they can also accept a singular event as their argument: one event can be trivially converted into a sequence of one element. Such ocorrence is so common and trivial that the conversion should therefore be implicit.

@@

One of the features that Python boasts (but are certainl not exclusive to it) that have eased our implementation are generators\cite{PEP255}. They integrate very nicely into both our concept of emitting musical events, as well as our concept of laziness where events that are not used are not generated either.

\paragraph*{Context State}
To keep track of the \textit{cursor} (the current timestamp where the next event should start) each operator in our language is implemented as a function call that receives an implicit \texttt{Context} object. While here we'll mostly focus just on the methods related to time management provided by the context, it can be used to store other types of information, like the default length of a musical note, to avoid forcing the user to type it out all the time.

@@

\paragraph*{Note Events}
The basic building block of our system is going to be a \textbf{Note Event}. We can then add even more events, like chords and rests, while the type of the event might change, the semantics are equivalent.

The function that emits the event receives the current context as an argument, and is then responsible for creating the event, with its timestamp matching the context's cursor. After the event is created, it also must move the context's cursor forward through the \texttt{seek} method.
\begin{lstlisting}[caption={Creating a Note Event},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
function note (ctx) {
    event = create_note(cursor(ctx));

    seek(ctx, event.duration);

    yield event;
}
\end{lstlisting}

\paragraph*{Concatenation}
We've seen how single events' creation is handled. Now it is important for us to see how we can combine those events together. And probably the most straightforward operator of all, concatenation, it simply consumes each event. Each event, as we've seen before, is responsible for seeking the context depending on the event's duration.
\begin{lstlisting}[caption={Algorithm to concatenate musical events},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
function concatenate (ctx, ...operands) {
    for (music in operands) {
        for (event in music(ctx)) {
            yield event;
        }
    }
}
\end{lstlisting}
\paragraph*{Repetition}
The repetition operand is in a way very similar to the concatenation operator. It makes sense, since repeating any kind of music pattern $N$ times could be thought as a particular case of as concatenation where there are $N$ operands, all representing the same musical pattern.

\begin{lstlisting}[caption={Algorithm for repetition},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
function repeat (ctx, music, count) {
    for (i = 0; i < count; i++) {
        for (event in music(ctx)) {
            yield event;
        }
    }
}
\end{lstlisting}

@@

\begin{lstlisting}[caption={Algorithm to merge parallel musical events},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
function parallel (ctx, ...operands) {
    for (child_ctx, event in merge_sorted(ctx, ...operands)) {
        join(ctx, child_ctx);

        yield event;
    }
}
\end{lstlisting}

@@

\subsection{Integration in a Programming Environment}
Apart from generating musical events from static instructions, our goal is to have those events integrate into a programming language in the same way integers, floats, strings and booleans do: as data that can be stored, passed around and manipulated. This, of course, while still retaining all the properties we've laid out for our sequences of events: being lazy and always being ordered.

\paragraph*{Variables}
Up until now we've analyzed situations where the timing of the events is known at the moment of their creation (through inspection of the context passed onto them). However, there can be situations where the events have to be created before their time location is known.

@@

>>>>

\section{Introduction}
Programming languages are sometimes described as data plus code. Some of that data is ingested through IO (like keyboards, files, sockets and others). Some of that data is generated by algorithms at runtime. But there's also a third category that blurs the line between data and code: literal values.

@@

When we cross into the territory of DSLs \textit{(Domain Specific Languages)}, there is an even bigger multitude of custom syntax for the most varied data types. In this paper we will discuss one possible way of of describing musical events (such as notes, chords or rests), and more importantly, musical arrangements and melodies. We will see how to treat these structures as data values that can be integrated into a programming language.

@@

The simplest way of generating such musical events is to use already common, \textit{low-level}, programming mechanisms, such as using a procedural approach where the user creates each event manually by calling a function and providing as parameters all the events' information, such as it's timestamp and duration. This is the approach used by some of the existing languages in this space, such as \textit{SonicPi}.

\begin{lstlisting}[caption={Example of a hypothetical imperative API for creating events},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
play_note( 0, 100, 'A' );
play_note( 100, 50, 'B' );
play_note( 150, 200, 'C' );
\end{lstlisting}

Instead we will take a look at a more \textit{functional} approach with custom syntax and operators. Musical events are treated as sequences, anc as such can be stored in variables, passed around and trasformed. For musical events, we will be exploring a way to define them in code, as \textit{musical literals}.

@@

The second requirement, \textbf{laziness}, refers to a concept that is familiar in functional programming languages: values are generated when we need them, not earlier. In our case, this implies that a musical sequence could be potentially infinite (like an infinite repetition of some arrangement). If playing this music live, the musician could determine to stop this arrangement sooner or later.

Given these two requirements, we can conclude we \textbf{cannot} generate all music events at the start and then sort them to play them in order. We conclude that the events must always be sorted already.

@@

%\begin{itemize}
% \item \textbf{NoteEvent} Describes a single note
% \item \textbf{ChordEvent} Describes a chord (group of notes to play in parallel). Is equivalent to having multiple separate \textbf{NoteEvent}'s with the same timestamp, but conveys a more specific semantic value.
% \item \textbf{ControlChangeEvent}
% \end{itemize}

@@

\paragraph*{Operators}
Operators are special operations defined at the syntactic level that allow \textit{music} to be composed in different ways, such as concatenated, parallelized or repeated. Many of these operators can have equivalent functions available through the language that provide more costumization (such as a parallel function that stops when the smaallest operand stops, instead of the longest).

\begin{description}
    \item[Concatenation] \verb|Music1 Music2 ... MusicN|
    \item[Parallel] \verb'Music1 | Music2 | ... | MusicN'
    \item[Repetition] \verb'Music * Integer'
    \item[Arpegio] \verb'Chord * Music'
    \item[Transpose] \verb'Music + Integer' \textit{and} \verb'Music - Integer'
\end{description}

It is also useful to estabilish that while most operators work on sequences of musical events, they can also accept a singular event as their argument: one event can be trivially converted into a sequence of one element. Such ocorrence is so common and trivial that the conversion should therefore be implicit whenever necessary.

@@

One of the features that Python boasts (but are certainl not exclusive to it) that have eased our implementation are generators\cite{PEP255}. They integrate very nicely into both our concept of emitting musical events as sequences (or iterators, as they are called in Python), as well as into our concept of laziness, where events are generated on demand when needed, and thus infinite musical sequences can be handled easily.

\paragraph*{Context State}
To keep track of the \textit{cursor} (the current timestamp where the next event should start) each operator in our language is implemented as a function call that receives an implicit \texttt{Context} object. While here we'll mostly focus just on the methods related to time management provided by the context, it can be used to store other types of information, like the default length of a musical note, for instance, to avoid forcing the user to type it out all the time.

It is important to keep in mind that there might be more than one context in execution at the same time. This can be most obvious with the use of the parallel operator, where each operand must run concurrently (and thus could not share the same context).

@@

\paragraph*{Basic Events}
The basic building block of our system are the \textbf{Note}, \textbf{Chord} and \textbf{Rest} events. We can use the current \textit{context} to determine the event's timestamp, as well as it's default duration (in case the user does not explicitly state one). Any event(s) that is/are not captured in a variable or passed to a function are implicitly played.

\begin{lstlisting}[caption={Creating a Note Event},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
c'1/4
\end{lstlisting}

\paragraph*{Concatenation}
We've seen how single events' creation is handled. Now it is important for us to see how we can combine those events together. And probably the most straightforward operator of all, concatenation, it simply consumes each event. Each event, as we've seen before, is responsible for seeking the context depending on the event's duration.
\begin{lstlisting}[caption={Snippet of Wet Hands by C418},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
S4/4 T74 L/8 V90;
A, E A B ^c B A E D ^F ^c e ^c A3;
\end{lstlisting}

\begin{figure}[h]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/concatenation2.png}}%
  }%
  \caption{Generated music sheet for concatenation\protect\footnotemark.}
  \label{fig:concatenation}
\end{figure}

\footnotetext{Rendered with \$ABC\_UI. Some hand made changes made for clarity.}

\paragraph*{Repetition}
The repetition operand is in a way very similar to the concatenation operator. It makes sense, since repeating any kind of music pattern $N$ times could be thought as a particular case of as concatenation where there are $N$ operands, all representing the same musical pattern.

\begin{lstlisting}[caption={Intro to Westworld's Theme by Ramin Djawadi},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
I1 S6/8 T140 L/8 V90;
A*11 G F*12
\end{lstlisting}

\begin{figure}[h]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/repetition2.png}}%
  }%
  \caption{Generated music sheet for repetition.}
  \label{fig:repetition}
\end{figure}

@@

\begin{lstlisting}[caption={Snippet of Soft to Be Strong by Marina},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
T120 V70 L1;
[^Cm] [BM] | r/4 ^g/4 ^g/4 ^f/2 e/8 ^d3/8 ^c
\end{lstlisting}

\begin{figure}[h]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/parallel.png}}%
  }%
  \caption{Generated music sheet for parallel}
  \label{fig:parallel}
\end{figure}

@@

\subsection{Integration in a Programming Environment}
Apart from generating musical events from somewhat static instructions, our goal is to have those events integrate into a programming language in the same way integers, floats, strings and booleans do: as data that can be stored, passed around and manipulated. This, of course, must still retain all the properties we've laid out for our sequences of events: being lazy and always being ordered.

\paragraph*{Variables and Functions}
Up until now we've analyzed situations where the timing of the events is known at the moment of their creation (through inspection of the context passed onto them). However, there can be situations where the events have to be created before their time location is known.

@@

Here we can see a small snippet of the beginning of Fugue 2 in C minor in Book I of the J.S. Bach’s Well-Tempered Clavier, and how using functions can help us visualize the structure behind music.

\begin{lstlisting}[caption={Example of repeating the same note},label=list:8-6,captionpos=t,abovecaptionskip=-\medskipamount]
fun fugue ( $subj, $resp ) => 
    ( $subj $resp | stretch( r, $subj ) ( $subj + 7 ) );

S8/4 T140 L/4 V120;

$subj = r c/2 B/2 c G _A c/2 B/2 c d
        G c/2 B/2 c d F/2 G/2 _A2 G/2 F/2;

$resp = E/2 c/2 B/2 A/2   G/2 F/2 _E/2 D/2   C _e d c
        _B A _B c   ^F G _A F;

play( fugue( $subj, $resp ) );
\end{lstlisting}

\begin{figure}[h]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/fugue.png}}%
  }%
  \caption{Generated music sheet for fugue example}
  \label{fig:parallel}
\end{figure}
% https://drive.google.com/open?id=1-igVwdhXMxOSE1cMg_8HimDOXMhsY9US
